---
title: "Untitled"
author: "ML"
date: "1/16/2020"
output: pdf_document
---

## Getting Started

Load the usual packages.

```{r, warning = FALSE, error = FALSE, message = FALSE}
library(tidyverse)
library(tidycensus)
library(pander)
```

So far we have used `tidycensus` to retrive data using "us", "state", and "county" as the geographic level. But the package allows for many other options as well. Here are some of them:

- region

- division

- tract (geometry shape files available)

- block group (geometry shape files available)

- place

- metropolitan statistical area/micropolitan statistical area

- combined statistical area

- urban area

- school district (elementary)

- school district (secondary)

- school district (unified)

- public use microdata area

- zip code tabulation area

- state legislative district (upper chamber)

- state legislative district (lower chamber)


## Metro Areas

As Manduca (2019) describes, metro areas are valuable geographic levels when analyzing inequality. Let's get the gini for all of them from the 2018 1-year ACS.

```{r}
gini_metro <- get_acs(geography = "metropolitan statistical area/micropolitan statistical area",
                       year = 2018,
                       survey = "acs1",
                       variables = c(gini = "B19083_001"))
```

Let's clean up the data. After changing the names to lower case, we want to keep only the metro areas (so get rid of the micro areas).

### REPLACE THIS LINE WITH YOUR CODE

```{r}
names(gini_metro) <- tolower(names(gini_metro))

gini_metro <- gini_metro %>%
  filter(!str_detect(name, "Micro Area"))
```


Now let's create a new data frame that only has the names and gini values for the twenty metro areas with the higest gini indices.

```{r}
top_gini <- gini_metro %>%
  select(name, estimate) %>%
  arrange(desc(estimate)) %>%
  top_n(20)
```

Create a new data frame that only has the names and gini values for the ten metro areas with the lowest gini indicies.

### REPLACE THIS LINE WITH YOUR CODE

```{r}
low_gini <- gini_metro %>%
  arrange(estimate) %>%
  select(name, estimate) %>%
  top_n(-10)
```

```{r}
pander(low_gini)
```



## Poverty

What variable do we need to calculate the poverty rate by state?

```{r}
poverty <- get_acs(geography = "state",
                   year = 2018,
                   survey = "acs5",
                   variables = c(total_pop = "B17001_001",
                                 total_pov = "B17001_002",
                                 white_pop = "B17001A_001",
                                 white_pov = "B17001A_002",
                                 black_pop = "B17001B_001",
                                 black_pov = "B17001B_002",
                                 aian_pop = "B17001C_001",
                                 aian_pov = "B17001C_002",
                                 asian_pop = "B17001D_001",
                                 asian_pov = "B17001D_002",
                                 nhpi_pop = "B17001E_001",
                                 nhpi_pov = "B17001E_002",
                                 other_pop = "B17001F_001",
                                 other_pov = "B17001F_002",
                                 twoplus_pop = "B17001G_001",
                                 twoplus_pov = "B17001G_002",
                                 whitenothisp_pop = "B17001H_001",
                                 whitenothisp_pov = "B17001H_002",
                                 hisp_pop = "B17001I_001",
                                 hisp_pov = "B17001I_002"),
                   geometry = TRUE,
                   shift_geo = TRUE)
```

Clean up the data

```{r}
names(poverty) <- tolower(names(poverty))
```

We can use the `separate()` funciton to break up our `variable` column into two separate columns. 

```{r}
poverty <- poverty %>%
  separate(variable, c("race", "number"), sep = c("_"))
```

Now calculate the poverty rate for each race in each state.

### REPLACE THIS LINE WITH YOUR CODE

```{r}
poverty <- poverty %>%
  group_by(name, race) %>%
  mutate(pov_rate = estimate / estimate[number=="pop"]) %>%
  filter(number=="pov") %>%
  filter(race != "total")
```

We know how to make one map. But we want a separate map for each race. In ggplot language, we want to "facet" the maps by the values of our race column.

```{r}
pov_race_map <- ggplot(poverty, aes(fill = pov_rate)) + 
  geom_sf() + coord_sf(datum = NA) + scale_fill_viridis_c() +
  facet_wrap(~race)

pov_race_map
```
