---
fontsize: 12pt
geometry: margin=1.0in
output: pdf_document
header-includes:
     - \usepackage{nopageno}
---

# Using Markdown For Reports

```{r, warning = FALSE, error = FALSE, message = FALSE}
library(tidyverse)
library(tidycensus)
library(pander)
library(mapview)

county_mobility <- read.csv("https://raw.githubusercontent.com/mjclawrence/soci1001/master/data/county_mobility.csv")
```


## Hiding Code and Inline Code
Let's start with a case where your output is a single summary statistic, like a mean. Imagine you are working on the descriptives part of your project and want to include the mean of absolute upward mobility. The place to start is with a regular code chunk with the `summary()` function:

```{r}
summary(county_mobility$absolute_upward_mobility)
```

If you just want the mean, you can use the `mean()` function:

```{r}
mean(county_mobility$absolute_upward_mobility, na.rm=TRUE)
# na.rm=TRUE removes the missing values
```

Say you want R to run a code chunk and have only the output - not the code! - show up in your knitted file. Simply add `echo = FALSE` to the first fence:

```{r, echo = FALSE}
mean(county_mobility$absolute_upward_mobility, na.rm=TRUE)
```

If you want to integrate a single number into your document, you can use inline code. Without opening a full code chunk, just use one backtick to open and close your fence. Then write a sentence as you normally would, and let R Markdown replace your code with the output:

The mean of age is `r round(mean(county_mobility$absolute_upward_mobility, na.rm=TRUE),2)`.


## Other Options For Hiding Code

If you want to run the code chunk so you can see the output in your notebook but with neither the code nor the output showing up in your knitted file, use `include = FALSE`.

```{r, include = FALSE}
mean(county_mobility$absolute_upward_mobility, na.rm=TRUE)
```

I would probably recommend starting with `include = FALSE` for your final project, so you can see all your output but then selectively choose what to include and what not to include in your knitted report.

If for some reason you want to show the code but not the output, use `eval = FALSE`.

```{r, eval = FALSE}
mean(county_mobility$absolute_upward_mobility, na.rm=TRUE)
```

## R Markdown Tips

Some other things to know about writing in R Markdown...

Use hashtags for headings. One hashtag is for a big heading; additional hashtags shrink the size. For example:

# Biggest Heading

## Big Heading

### Small Heading

#### Smallest Heading

If you want to italicize text, *wrap it within single asterisks*. If you want to bold text, **wrap it within double asterisks**. And if you want to italicize *and* bold text, ***wrap it within triple asterisks***.

It can sometimes be helpful to highlight original variable names or unusual terms within tickmarks. But note this is similar to the inline code we saw earlier. As long as the word or phrase does not start with a single r, R will not try to run it as code. See the preview file for the difference in what these tickmarks represent:

The mean of `absolute_upward_mobility` is `r mean(county_mobility$absolute_upward_mobility, na.rm=TRUE)`.

To create an ordered list, leave an empty line and then:

- Start
- Each
- Item
- With 
- A
- Dash

To create a numbered list, leave an empty line and then:

1. Start
2. Each
3. Item
4. With
5. A
6. Number and a period

To add a horizontal line rule, include at least three dashes on a single line:

---

And to add a page break:

\pagebreak

This should be the start of a new page!

# It's Also The Start Of A New Section