---
title: "Introducing R"
author: "ML"
date: "1/9/2020"
output: pdf_document
---

## Getting Started

To begin, copy all the text on this screen. Then open RStudio, and in the "File" menu select "New File" then "R Markdown". Add a title like "SOCI 1001, Week One, Class Three", include your name in the "Author" box, select "PDF" as the Default Output Format, and click Ok. A default markdown file will open. Delete all the text in the default file below the header (everything after line 7) and paste in the text you copied to your clipboard. Make sure that "R Markdown" is selected in the file type drop down menu in the bottom right corner of this pane. Save this file to your desktop.

## Welcome to R Studio!

Some navigation before we begin:

- The upper left pane is the *text editor* where you keep all of your commands.
- The lower left pane is the *console* where the commands will be executed.
- The upper right pane's `Environment` window will show the data that you load.
- The lower right pane will show previews of your R Markdown documents, plots, help documents, packages, and other features.

This file is an R Notebook which is an R Markdown file that allows you to integrate text, R code, and R output into a single document. This is the recommended file type for our course, but you may want to use other file types in the future. All the possible file types are found when you click File>New File. For our purposes - and for producing transparent and reproducible research - the advantage of an R Notebook is that you can easily include text, notes, and comments that are not code (like this paragraph).

How does R know what is code and what is regular text? In an R Notebook, code goes in a fenced code block. To open the fence, type three backticks and an r in curly brackets on a single line; to close the fence, type three backticks on a single line like this:

```{r}

```

Everything between those fences will be run as code (unless it is preceded by a hashtag).

Let's start by using R as a basic calculator. Since even the following calculation is considered code we need to place it within a code block. After the r, in the line where you open your fence, you can provide a short description of what the code in that chunk does. Those short descriptions are collected as bookmarks in the drop down menu on the bottom of this pane and can help you find pieces of your file.

```{r calculator 1}
1000 + 1
```

To run this line of code, click the green arrow on the far right side of the first line of the code block.

For subtraction, use a -; for multiplication, use a *; for division, use a /. 



What is 2002 divided by 2?

### REPLACE THIS LINE WITH YOUR CODE ###

```{r calculator 2}
2002 / 2
```

Our course number could be helpful to save so we do not have to calculate it each time we want to reference it. We save a value by creating an object. The syntax for doing so is to name our object first, then use the symbol "<-" to say what value our object takes. Let's call our new object `coursenumber`:

```{r creating an object}
coursenumber <- 2002 / 2
```

Run this line of code. In the top right quadrant, we should see a new value saved. That's promising, but there's no output below the code chunk! That's because we have to separately recall the object for it to be displayed. The only line of code we need to do so is our object name:

```{r recalling an object}
coursenumber
```

Note that case matters in R code. See how by replacing the `c` in the previous chunk with a capital `C`.

We can now use the object we created in other calculations by referring to it by name:

```{r calculator with object}
coursenumber - 145
```

Objects can take numerical or character values. The difference is that if we are using text we have to enclose the value in quotation marks:

```{r object with characters}
coursename <- "Counting The US: The 2020 Census"
coursename
```

We can now reference these saved objects in our code as well as our text. To use *inline code*, wrap the code in single backticks with a preceding r:

> This class is called `r coursename` and it is SOCI `r coursenumber`.

The printed document will replace `coursename` and `coursenumber` with their saved values.

## Introducing R Markdown

We will be using R Notebooks with R Markdown to write and run our code. R Markdown is a plain text-based form of writing with lots of neat tricks. For example, look at the heading for this paragraph. The two hashtags are how we start a new section in R Markdown. If we used one hashtag, the section heading would be bigger in the output. If we used three (or four) hashtags, the section headings would be smaller (or even smaller) in the output, which would be helpful if you have lots of sub- (or sub-sub-) headings

There's an R Markdown cheat sheet in the pages folder on Canvas that has much more info on how to use R Markdown. Here are a few first day features to get you started:

Wrapping a word in `single backticks` will highlight it when you print.

*Wrapping text in one asterisk will italicize it when you print.*

**Wrapping text in two asterisks will bold it when you print.**

***Wrapping text in three asterisks will bold and italicize it when you print.***

And three (or more) dashes on a single line will create a horizontal rule:

---

\pagebreak

## Getting Started With R

Today we will learn one way to import data into R and how to get simple summaries of different types of variables.

We will use the Census Bureau's 2018 population estimates for every county to explore types of variables.

First, we need to install a package. Packages are additional functions that supplement base R. We will use the `tidyverse` package extensively this semester.

Here's the code to install the package. You only have to install the package once on any computer you use. Once your computer has the package installed, you can load the package with the `library()` function.

```{r load packages, warning=FALSE, error=FALSE, include=FALSE}
#install.packages("tidyverse")
library(tidyverse)
```

Now that the package is installed, you can put a hashtag in front of the `install.packages()` line so it does not run again.

Next we'll get the data.

```{r}
week_1_3 <- read.csv("https://raw.githubusercontent.com/mjclawrence/soci1001/master/data/week_01_03.csv")
```

If the code above executed correctly, you should now see the data frame loaded and saved with the object name you created in the `Environment` panel in the top right quadrant of R Studio. There should be 3142 observations of 16 variables in the data frame named hurricanes.

Click the white arrow to the left of the object name to see all the variables (geoid, county, state, etc.). Click the spreadsheet icon on the far right side of the data name row to open the full data frame. Note that the *variables* are in the columns and the *values* are in the rows, with each row representating a separate *observation* or *case*.

## Types of Variables

We have different types of variables in our data frame. Some have the `int` tag identifying them as interval variables, some have the `Factor` tag identifying them as categorical variables, and some have the `numeric` tag.

Let's look at the `state` variable to start. We can see in the `Environment` panel that this variable is a factor variable with 51 levels - or 51 different possible values. In this case, each value is a different name, making this an example of a *nominal variable*.

To see how many observations have each value - or how many counties are in each state - we will use the `table()` command. In R, we have to reference all variables by their data frame. So if we want a table of the `name` variable, we first have to tell R that the variable is in the `week_1_3` data frame. We separate the data frame and the variable name by a dollar sign:

```{r table of factor variable}
table(week_1_3$state)
```

When you run that line of code you should see a table with every possible value of the `state` variable and the number of observations with that value. There are 67 counties in Alabama, for example. How many counties are in Vermont? Which state has the most counties? 

By default, R will sort factor variables alphabetically. An easier way to see which factor has the most observations - in this case, which state has the most counties - is to use the `sort()` function to order the names by the number of observations in each. Note that functions "wrap" code in parentheses:

```{r sort table}
sort(table(week_1_3$state))
```

To sort the table from highest to lowest value, add the `decreasing = TRUE` option:

```{r sort table with decreasing}
sort(table(week_1_3$state), decreasing=TRUE)
```


Now let's look at the numeric variable `rnaturalinc`. This variable is the *rate of natural increase*. Positive values tell us there are more births than deaths per 1000 residents; negative values tell us there are more deaths than births per 1000 residents.

To see the distribution of a numeric variable, we use `summary()` instead of `table()`.

```{r}
summary(week_1_3$rnaturalinc)
```



